---
title: "Plotting"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_float: false
    toc_depth: 4
    collapsed: true
vignette: >
  %\VignetteIndexEntry{Notes 11: Plotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center',
  fig.height = 3, fig.width = 3
)
```

![plotting][plottingcat]

## Plot

- R has rich graphics capabilities built-in, and even more are added by external packages
- the graphical environment is constantly growing, just like the computational side of R
- base R graphics is implemented in the `graphics` and `grDevices` packages
- `graphics` contains a number of functions for creating and formatting plots

<br>

#### creating a plot

- the basic plotting function is `plot`, which (usually) draws scatter plots
- the first thing it does is to craete a new plot (with an internal call to `plot.new`)
- it then calls an appropriate plotting method
- plot has a number of methods and can accept various input formats

```{r}
# a single numeric vector
# elements are plotted against their indices
v <- 1:10
plot(1:10)
# two numeric vectors
vs <- sample(v)
plot(x = v, y = vs)
# a matrix
# the first two columns are plotted as if they were x and y, the others are ignored
m <- matrix(1:12, 4, 3)
m
plot(m)
# a data frame
# every column is plotted against every other, creating a matrix of scatter plots
d <- iris[3:5]
plot(d)
```

- note that axes are labelled automaticaly and informatively, if not beautifully
- also note that scatter plots render factors as integers

<br>

#### plot content

- plot appearance is handled by changing **graphical parameters**
- this is done with the `par` function, which is called internally by `plot`
- arguments to `par` are passed through `...`
- see `?par` for a full list of available parameters; here are some examples:

\newline

- `type` decides the plot type, e.g. points, lines or both
```{r}
plot(v, vs, type = 'p')
plot(v, vs, type = 'l')
plot(v, vs, type = 'b')
```

<br>

- `pch` sets the plotting character for points
- available values are 1 through 25
```{r, fig.width = 6}
plot(1:25, rep(1, 25), pch = 1:25)
```

- note that assigning `pch` is a vectorized operation
- above the length of `pch` was the same as that of `x` but `pch` will be recycled if necessary
```{r, fig.width = 6}
plot(1:25, rep(1, 25), pch = 1:5)
```

- this behavior is common to many graphical parameters

<br>

- `lty` controls line type
```{r}
plot(v, vs, type = 'l', lty = 1)
plot(v, vs, type = 'l', lty = 2)
plot(v, vs, type = 'l', lty = 3)
```

<br>

- colors are handled with `col`
- `col` can be specified in a number of ways:
```{r}
# as a hex code
plot(v, vs, pch = 16, col = "#6495ED")
# as a character string
# see colors() for a list of available colors
plot(v, vs, pch = 16, col = "cornflowerblue")
# as a factor; each level is given a color based on the current palette
f <- factor(rep(letters[1:2], each = 5))
plot(v, vs, pch = 16, col = f)
# as an integer, which is interpreted as an index in the current palette
plot(v, vs, pch = 16, col = 2)
```

- palettes are character vectors of hex codes or color names
    - they can be created manually
    - or with predefined functions like `rainbow`
    - color gradients can be created with `colorRampPalette`
    - get or set the current palette with `palette`
    
*NOTE: `colorRampPalette` is a function that returns a function which takes an integer as argument and returns a vector of that many interpolated colors.*
```{r}
plot(v, vs, pch = 16, col = colorRampPalette(c('royalblue1', 'yellow', 'purple'))(10))
```

<br>

- `xlim` and `ylim` set the axes limits
```{r}
plot(v, vs, xlim = c(-2, 12))
```

<br>

#### adding content

- more data can be drawn on an existing plot with `points` or `lines`
```{r}
plot(v, vs)
points(v, sample(v), pch = 20, col = 'green')
```

- they work nearly the same as `plot`
- they even respond to the `type` parameter, so you can draw lines with `points`
- since they do not create a new plot, some data can fall outside of the window

<br>

- `abline` adds a straight line to an existing plot
- arguments `a` and `b` describe its intercept and slope, respectively (slope is the tangent of the angle between the line and the X axis)
- arguments `v` and `h` draw vertical and horizontal lines, respectively, and determine their crossing points with the respective axes
```{r}
plot(log(0:2000 / 100), type = 'l')
abline(h = 0)
plot(log(0:2000 / 100), type = 'l')
abline(h = c(0,1), lty = c(1,2))
```

<br>

#### plot appearance

- the size of elements is controlled by the `cex` parameter
- when a new plot is created, its size is determined by the size of the current device (see below)
- the default size of all plot elements is derived from that plto size
- `cex` is the multiplier to that default size
```{r}
plot(v, vs, pch = 16)
# cex alone refers to the plotting characters
plot(v, vs, pch = 16, cex = 2)
# other cex parameters exist for other plot elements
plot(v, vs, pch = 16, cex.lab = 2)
```

<br>

- `xlab` and `ylab` change axis labels
```{r}
plot(v, vs, xlab = 'label x', ylab = 'label y')
```

<br>

- `main` sets the plot title
```{r}
plot(v, vs, main = 'plot title')
```

- the title of an existing plot can be set with `title`
```{r}
plot(v, vs)
title('plot title')
```

<br>

- `legend` adds a legend box
```{r}
plot(v, vs, col = f, pch = 16)
title('plot title')
legend('topright', legend = levels(f), col = 1:length(levels(f)), pch = 16, cex = 0.75)
```

<br>

- `las` controls the orientation of axis labels

<br>

#### plotting area

- the plotting area, by default, fits only one plot
- this can be changed with the functions `mfrow`, `mfpar` and `layout`, of which I only use the latter
- `layout` takes a matrix of integers as argument `mat`
- the elements of `mat` are 1 through the number of plots to be drawn
- the plotting window is divided into `length(mat)` fields
- plots are put into fields based on the values of corresponding elements of `mat`
- after the last field has been drawn in, the next `plot` call will create a new window with the same layout
```{r}
m <- matrix(c(1, 1, 2, 1, 1, 2, 3, 3, 4), 3, 3, byrow = T)
m

layout(mat = m)
```

- `layout.show(n)` shows which field in the current layout the *n*-th plot will appear in

- here is an example (wrapped in a function to work within the vignette):
```{r, fig.height=6, fig.width=6}
foo <- function() {
  # prepare layout
  layout(mat = m)

  # prepare data
  v1 <- rnorm(1e4)
  v2 <- runif(1e4, -1, 1)
  d1 <- density(v1)
  d2 <- density(v2)
  names(d2) <- c(names(d2)[2], names(d2)[-2])
  
  # draw plots
  plot(v1, v2, pch = 4, cex = 0.5)
  plot(d2, main = '', xlab = '', ylab = '')
  plot(d1, main = '', xlab = '', ylab = '')
  # draw an empty plot in field 16
  plot.new()
}

foo()
```

- note that plots 2 and 3 are created by the method for class density
- also note that each plot is entirely separate and could have its own title, legend, etc.

\newline

- resetting the layout is easy
```{r}
layout(mat = 1)
```


<br><br><br>


## Other Plot Types

- there are many built-in plotting functions other than `plot`
- listing them all would be silly but let's go through some

<br>

- scatter plot
    - this is what `plot` does
    - large samples are drawn prettier with `smoothScatter`
```{r}
plot(rnorm(1e4), runif(1e4), pch = 4, cex = 0.2)
smoothScatter(rnorm(1e4), runif(1e4), pch = 4, cex = 0.2)
```

<br>

- `barplot`
    - draws a typical barplot
    - the vector method draws one bar per element
    - the matrix method draws one bar per element but groups the bars according to columns
    - by default bars within columns are stacked; they can be separated by setting `beside = TRUE`
    - calling `plot` on a factor will draw a barplot
```{r}
# a vector
barplot(v)
# a matrix
barplot(m)
barplot(m, beside = T)
```

<br>

- 1d scatter plot
    - sapmles/categories are drawn separately
    - all points of a sample are drawn such that they do not overplot
    - implemented with the function `stripchart`
```{r}
# a vector
rv <- rnorm(1e2)
stripchart(rv, pch = 20)
stripchart(rv, method = 'jitter', pch = 20)

# a data frame
stripchart(iris[1:4], method = 'jitter', vertical = T, pch = 1)
```

*NOTE: This is implemented better in package `beeswarm`.*

<br>    

- boxplot
    - very useful type of plot
    - highly recommended over barplots
    - starts working as sample size of about 7
    - draws a graphical representation of basic statistics of the sample
    - a box delimits the interquartile range
    - the median is marked with a horizontal bar
    - the range of the data is marked by lines and "whiskers"
    - outlying data points are highlighted
    
    - boxplots are drawn with `boxplot`
    - the `varwidth` argument makes the box width proportional to sample size
    - the `notch` creates indentations in boxes: if these do **not** overlap, the means are "significantly different"
```{r}
# a vector
boxplot(rnorm(1e3))
# a matrix
boxplot(as.matrix(iris[1:4]))
#a data frame
boxplot(iris[1:4])
```

<br>

- histogram
    - a plot of observed frequencies of events
    - extremely useful for assessing a sample's distribution
    - implemented with the `hist` function
    - it bins a numeric vector and draws frequencies of the bins
    - the `breaks` argument controls the number of bins
    - unlike most plotting funcionts, `hist` also returns the plot as an object that can be drawn alter with `print`
```{r}
hist(rnorm(1e5))
hist(rnorm(1e5), breaks = 100)
```

- plotting 3d data
    - built-in tools for 3d data are fairly limited
    - `persp` draws a surface
    - `contour` draws contours
    - `image` draws a color-coded matrix
```{r}
persp(volcano)
contour(volcano)
image(volcano)
```


<br><br><br>


## Plotting With Formulas

- many plotting functions can be driven by a formula
- the formula is evaluated within a data frame or a list given by the argument `data`
- if `data` is undefined, the formula will be evaluated in the Global Environment
- formula methods usually have the `subset` argument
    - it accepts logical predicates to select observations from `data`
    - it uses non-standard evaluation
    - the documentation for the `subset` function can be informative here
- the formula interface of `plot` can also evaluate the `col` argument with NSE

<br>

#### formulas in `plot`
```{r}
# numeric vs numeric draws a scatter plot
plot(Sepal.Length ~ Sepal.Width, data = iris, col = Species, pch = 20)
# numeric vs factor drawas a boxplot
plot(Sepal.Length ~ Species, data = iris, col = Species)
# two terms on RHS draws two plots
plot(Sepal.Length ~ Petal.Length + Petal.Width, data = iris)
# plot one column vs all others
plot(Sepal.Length ~ ., data = iris, col = Species, pch = 20)

```

<br>

#### formulas in other plotting functions
```{r}
barplot(yield ~ variety + year, data = lattice::barley, subset = site == 'Morris', beside = TRUE)
stripchart(Sepal.Length ~ Species, data = iris, vertical = T, method = 'jitter', pch = 1)
boxplot(Sepal.Length ~ Species, data = iris)
```

<br>

#### enhanced use of formulas

- `coplot` uses more complex formulas to separate data into subplots
```{r}
coplot(Sepal.Length ~ Sepal.Width | Species, data = iris, pch = 16, rows = 1)
```

- this is also, quite robustly, implemented in the `lattice` package
```{r, fig.width=6, fig.height=3}
library(lattice)
xyplot(Sepal.Length ~ Petal.Width | Species, data = iris, layout = c(3,1))
stripplot(yield ~ year | site, data = barley, layout = c(3,2))
bwplot(yield ~ year | variety, data = barley, layout = c(5,2))
bwplot(Sepal.Length ~ Species, data = iris)
histogram(~ Sepal.Length | Species, data = iris, layout = c(3,1))
levelplot(Sepal.Length ~ Petal.Width * Petal.Length | Species, data = datasets::iris, layout = c(3,1),
          at = seq(4, 8, length.out = 100), col.regions = colorRampPalette(c('blue', 'gray90', 'red'))(100),
          aspect = 1)
```

*NOTE: `lattice` plots ignore some `par` commands, e.g. `title`, `layout`.*

*ALSO NOTE: `lattice` plots are only drawn automaticaly when in an interactive session. Otherwise they must be drawn explicitly with `print`.*


<br><br><br>


## Saving Plots To Files

The simplest way to save a plot to a file is to draw it in RStudio viewer and press the "Export file" button in the viewer but where is the fun in that? You will also find the viewer rather slow in some cases.

#### graphics devices

R has a system of **devices** that graphics is streamed to. The default device is the **null device**. Graphics sent to the null device ends up in the RStudio viewer in the Plots tab. Other devices can be opened at any time to send the graphics to a file. Opening a new device creates a file to capture the incoming graphics. That device becomes the current device, where the graphics will be sent now. Functions that open devices include `jpeg`, `png` and `pdf`, each for the corresponing file type.

All currently open devices can be viewed with `dev.list()`. A newly open device is added to the end of the list and becomes the current one. It can be viewed with `dev.cur()`. There are ways to switch between devices but that is rarely done. Usually one opens a device, streams the graphics and closes the device immediately after.

Devices are closed with the `dev.off` function. It will close the device specified by a number, that device's number on the list. Called without an argument, `dev.off()`, it will close the current device. **Only when a device is closed is the corresponding file saved.** Attempts to manipulate the file before that will throw errors. Opening a stream to a file that is already associated with a graphics device, which can happen when plotting fails and one wants to immediately try again, is a sure way to trouble. Sometimes a `dev.off()` needs to be redoubled.

`graphics.off` closes all open devices. The null device can never be closed. `graphics.off()` skips it and `dev.off(1)` throws an error. Calling `dev.off()` on a null device will also clear the plots from RStudio memory.

<br>

#### printing plots

To print a plot to a file open a device with, say, `png`, draw a plot as you normally would in RStudio, and call `dev.off()` to close the file. Unlike in RStudio, past plots are not available for rewinding. Only the last plot is saved to the file. The `pdf` device behaves differently. It has an `onefile` argument, which is set to `TRUE` by default. With the pdf device plots will be added to the file, one plot per page. Again, the graphics cannot be viewed before the device is closed.

Also, .pdf files are saved as vetor graphics rather than rasters. You can open them in Corel or Illustrator and edit partilcular elements there with no loss of quality.

[plottingcat]: ../images/plottingcat.png
