---
title: "Notes For Part Two"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Notes For Part Two}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Data Structures And Their Properties

## Vector
- a **vector** is a sequence of items in a certain order
- it is a 1-dimensional structure of any **length** (number of items)
- a vector is the simplest data structure in R, hence it is called **atomic**
- virtually every data object is some kind of vector
- even a single data point, e.g. a `TRUE`, is actually a vector of length 1
- a vector can store only one type of data, e.g. logical or character, thus it has a type
- vectors are created with the function `vector`: it can create any type (mode) of vector of any length
- constructors for particular types exist: `logical`, `numeric`, `character`
- functions exist for testing and converting types: `is.character`, `as.character`, etc.

\newline

- vectors are most commonly built by concatenating/merging vectors with the function `c`
```{r}
# merge two character vectors of length one
c('one', 'two')
```
- types will often be changed to suit a called function: this is called **coercion**
```{r}
# merge a character vector with a numeric one
c('one', 1)
# the numeric was (silently!) coerced to character
```

```{r}
# concatenate existing vectors (variables) with one created ad hoc
v1 <- c('one', 'two')
v2 <- c(1, 2)
v3 <- c(v1, v2, c(FALSE, TRUE))
v3
```
- a vector can be treated as a set
- use `%in%` to test if something is an element of a vector:
```{r}
'one' %in% v3
```
- a vector can contain duplicated elements
- use `anyDuplicated` to test if that is the case
- use `unique` to get unique elements of a vector

\newline

- a vector can have additional features, such as properties or information
- these are called **attributes**
- all attributes of an object can be viewed with the function `attributes`
- a specific attribute can be viewed with `attr`
- a frequently encountered attribute is **names**: unique character strings associated with each item
- a vector that has a names attribute is a **named vecotr**
- see `.romans` for a named vector of Roman numerals
- names can be viewed with `names(x)` and set with `names(x) <- [vector of names]`, much like assigning a variable
- this would be equivalent to `attr(x, 'names')`; `names` exists for convenience, as messing with names is such a common action
```{r}
# get names
names(v2)

#set names
names(v2) <- c('first', 'second')
names(v2)

# remove names
names(v2) <- NULL
names(v2)
```
- a vector can also be created with names when using `c`
- elements are given as `name = value` pairs
```{r}
v4 <- c('first' = 1, 'second' = 2, 'third' = 3)
names(v4)

# the names need not be quoted
v4 <- c(first = 1, second = 2, third = 3)
```

- we will discuss more attributes when appropriate but names will keep coming back
- the function `as.vector` will strip all atributes, leaving a bare vector

#### Every vector has a type and a length. It can also have attributes.



## Vectorized Operations

## Indexing And Subsetting


## Matrix
- a matrix is a 2-dimensional data structure
- it has rows (dimension 1) and columns (dimension 2)
- a matrix is created with the function `matrix`
```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)
m
```
- by default a matrix is filled **by columns**, i.e. 1st column from 1st to last row, then 2nd column, etc.; this can be modified wyth the `byrow` argument of `matrix`
- a matrx can be transposed with the funcion `t`
```{r}
t(m)
```
- we can check the number of rows and columns with `nrow` and `ncol`, respectively

\newline

- a matrix is not a special atomic type with unique properties
- it is merely a vector with an attribute **dim**
- dim can be viewed or set with a special function `dim`
```{r}
# get dimensions
dim(m)
# the first number is the number of rows and the second is the number of columns

# the attribute can be modified as will, even removed
dim(m) <- NULL
m
# note that the matrix is dismantled in the same way is is built by default:
# column after column

# dim can be set on any vector as long as the grid size matches its length
dim(m) <- c(4,3)
m
# note that, again, the matrix is build in the default order
```
- since a matrix is basically a vector, it can hold only one type of data
- it also has a length:
```{r}
length(m)
```

#### Subsetting a matrix
- when subsetting a matrix, we must account for it having two dimensions
- specify the range of each dimension and separate them with a comma:
```{r}
m[1,3]
m[1:2, 2:3]
```
- leaving one index empty will yield a whole row/column
- be aware this returns a vector, not a 1 row matrix!
```{r}
m1 <- m[1, ]
is.matrix(m1)
is.vector(m1)
```
- in order to preserve tha matrix nature of the result, we must set the `drop` argument to FALSE:
```{r}
m2 <- m[1, , drop = FALSE]
# mind the empty second dimension!
```
- since it is a vector, we can also subset using a single index, bypassing the dimensions:
```{r}
m[1]
m[7]
```

\newline

- a matrix can be extended by adding rows or columns with `rbind` and `cbind`, respectively:
```{r}
rbind(m, 101:103)
cbind(m, letters[1:4])
# note the coercion
```
- `c` would do no good here as it strips attributes other than names
```{r}
c(m, m)
cbind(m,m)
```

\newline

- a matrix's dimensions can be named:
```{r}
rownames(m) <- letters[1:4]
colnames(m) <- LETTERS[1:3]
m
```
- we can now use the names for subsetting
```{r}
m['a', 'B']
m['a', c('A', 'B'), drop = FALSE]
m[c('a', 'b'), c('A', 'B')]
```
- the row and column names are stored a single attribute **dimnames**
- dimanmes can be viewed or set with the function `dimnames`
- it is structured as a list (more on lists below)



## List

## Data Frame







```{r ntoes}
# ###### vector functions
# - `c`: build a vector from existing elements
# - `vector`, `character`, etc.: create vectos with default values (e.g. 0)
# - `is.character`, etc: test vector type
# - `length`: get length (number of items) of a vector
# - `%in%`/`is.element`: test if an item is an element of a vector
# - `anyDuplicated`: test whether vector contains duplicates
# - `unique` get unique elements of a vector
```

