---
title: "Notes For Part One"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Notes For Part One}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## I RStudio basics:
-	the Console is on the bottom left and that is where the magic happens
-	the upper left is the Source, or script, from where we can execute consecutive lines of code
-	the upper right shows the contents of the Global Environment, which can be useful
-	help and plots will appear on the bottom right

\newline

-	use Ctrl+1 to go to Source and Ctrl+2 to go to Console (and Ctrl+3 to go to the help tab)
-	when in source, use Ctrl+Enter to execute the current line of code
-	use Ctrl+Enter to execute a highlighted block of code: several lines or part of a line

\newline

-	workspace and console history can be saved and loaded between sessions; I myself never do it

## II Console basics:
-	execute a statement with Enter
-	use Up and Down arrows to navigate console history
-	use Ctrl+L to clear the console (does not affect history)
-	call for help like this:
```{r help, eval=FALSE, include=TRUE}
?Rdlazer
help(Rdlazer)
help("Rdlazer")
```

-	each statement is executed separately
-	by convention each statement is put in a single separate line
-	each line may, but need not, end with a semicolon (;)
-	a single line can hold multiple statements, provided they are separated by semicolons
-	a single statement can span multiple lines; in some cases, this can improve intelligibility
-	a block of code contained within braces will be executed in its entirety
-	when the cursor is on or after a close\* sign, the appropriate open\* sign is highlighted
-	this goes for **parentheses** ( **(** ), **brackets** ( **[** ) and **braces** ( **{** )

## III comments
- the pound sign (**#**) is a comment sign
- everything right of a # in a line is considered a comment and not code and will not executed
- comments are put in source rather than the console; you want the console to do things, after all
- commenting code is very helpful for other people to understand code
- other people include yourself some (surprisingly short) time in the future



\newpage
## IV basic programming from an R perspective:
-	there are four basic **types** of data: **logical**, **character**, **integer** and **double**
-	there are other types in R but that is deep lore and we don’t need it here
-	integer and double are much the same for most purposes, so they are both treated as **numeric**

####character strings
-	character strings are delineated by **quote marks**, single or double
-	to place a quote mark in a string you must **escape** it by preceding it with **\\**
-	whitespace can be encoded in strings with **special characters**: **\\t** for tab, **\\n** for new line
-	in these cases the backslash bestows special meaning rather than removes it as with quotes

<h4> Boolean logic <br> 
<font size = "1"> 
*named after British mathematician George Bool; it is said he died because he was set to FALSE* 
</font></h4>

-	logical values can be **TRUE** or **FALSE**; these are reserved words (you cannot use them as names)
-	**T** and **F** are aliases for TRUE and FALSE; these are not reserved
-	logical values can be interpreted as integers with 1 for TRUE and 0 for FALSE
-	logical operators are:
    - **==**	&nbsp;&nbsp;&nbsp;&nbsp; test identity; is this the same as that?
    - **!**		&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; negate; reverse a logical value
    - **!=**	&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; test difference; is this different to that?
    - **&**   &nbsp;&nbsp;&nbsp;&nbsp;&nbsp; AND; are both items TRUE; analogous to multiplication
    - **|**		&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; OR; is any of the two items TRUE; analogous to addition

####simple maths
-	most arithmetic operations are implemented as operators:
    - **+**, **-**, **\***, **/**, **^**	&nbsp;&nbsp;&nbsp;&nbsp; add, subtract, multiply, divide, raise to power
    -	**%%**, **%/%**		                  &nbsp;&nbsp;&nbsp;&nbsp; division remainder, modulus
-	numeric values can be compared by operators:
    -	**==**, **!=**  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; test equality/inequality
    -	**>**, **<**, **>=**, **<=**  &nbsp; greater than, less than, greater or equal, less or equal
    -	operation precedence is as usual, including parentheses

####variables
-	a **variable** is a **binding** of a **value** to a **name** (a.k.a. **symbol**)
-	a variable is **created** or **declared** by **assigning** a value to a name
-	a variable, as well as a binding, exists in an **environment** (more on environments later)
-	variables can be created by **<-**, **->** and **=** (name on left and value on right)
-	it is good style to always use **<-**, with spaces around it for intelligibility
-	**<<-** and **->>** also assign variables but do not use them unless you have to (more on this later)
-	when creating a variable, its type is determined automatically (unlike in, say, Java or C)
-	a new value can be assigned to an existing symbol, overwriting the previous value

\newpage
####functions
-	where a variable is a thing that ***is***, a function is a thing that ***does***
-	a function is a block of code that can and will perform a sequence of operations
-	some functions consist of R code but many are written in C for efficiency and cannot be read
-	functions are stored in memory as a special type of object called **closure**
-	a function will usually take input, possibly some existing variables/objects
-	a variable that is used by a function is said to be **passed** to a function as an argument, see below
-	variables passed to a function remain unaffected
-	a function will usually produce output
-	the output value is **returned**
-	an instance of a function’s action is a **function call**
-	a function’s behavior can be modified; this is done by setting **arguments**, a.k.a. **parameters**
-	a function is **called** by writing its name followed by parentheses containing arguments: `foo(x, y)`
-	a function’s definition contains a list of arguments in a particular order
-	some arguments have default values (given in the definition) and can be omitted from a call
-	there are three ways of matching a value to an argument:
    -	by name: arguments in a call are provided as *name = value* pairs
    -	by partial name: the name need not be complete as long as it can be matched uniquely
    -	by position: unnamed values are matched in the order they appear in the function definition
-	again, passing a variable as an argument does not alter that variable
-	some functions work as **infix operators**: the call is structured `x op y` rather than `foo(x, y)`
-	examples of this are: `%in%` or `+` or `<-`

\newpage
####flow control
several special constructs allow for controlling code behavior

######1. perform actions conditionally with **if** and **else**:

-	the **if statement** consists of a condition and an expression:

```{r if, eval=FALSE, include=TRUE}
if (condition) {
  expression
}

```

- **expression** (the block of code within the braces) is only executed if condition is TRUE

\newline

- if statements can be nested:
```{r if-nest, eval=FALSE, include=TRUE}
if (condition1) {
  if (condition2) {
    expression1
  } else {
    expression2
  } else {
    expression3
  }
}
```
- expression1 will be executed only if both conditions are TRUE
- condition2 is tested only if condition1 is TRUE

\newline

-	an if statement can be followed by an **else statement**, which contains an alternative expression:
```{r if-else, eval=FALSE, include=TRUE}
if (condition) {
  expression1
} else {
  expression2
}

```
-	the alternative expression will only be executed if the condition is FALSE

\newline

-	else and if can be used to build a sequence of alternatives with multiple conditions:
```{r if-else-mult, eval=FALSE, include=TRUE}
if (condition1) {
  exrpession1
} else if (condition2) {
  exrpession2
} else if (condition3) {
  exrpession3
} else {
  exrpession4
}

```

-	after one condition is found to be TRUE, the others are not tested
-	only the first block of code for which the condition is TRUE will be executed
-	*WARNING: never brake a line before an else!*

\newline

######2. perform actions iteratively with **loops**:

-	a loop will execute a set of expressions (the **body**) iteratively
-	the **next** command will interrupt an iteration and move the loop on to the next one
-	the **break** command will terminate the loop immediately

\newline

-	a **for** loop: 		
```{r for, eval=FALSE, include=TRUE}
for (iterator in sequence) {
  expression1
  expression2
  expression...
  }
```

\newline

-	the **iterator** takes on the values of consecutive items in the sequence in each iterations
-	the expressions in the body of the loop can make use of the iterator, e.g. pass it to functions

-	a **repeat** loop: 	
```{r repeat, eval=FALSE, include=TRUE}
repeat {
  expression1
  expression2
  expression...
}
```

-	the body of the loop will be repeated over and over until a break is encountered

\newline

-	a **while** loop:	
```{r while, eval=FALSE, include=TRUE}
while(condition) {
  expression1
  expression2
  expression...
}
```
-	the condition is tested before every iteration and the loop runs as long as it is TRUE
