---
title: "Notes 02: <br> Data Structures And Language Properties"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Structures And Language Properties}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(Rdlazer)
```

![a tour of R][confusedcat]


<br><br><br>


## I Vector

- a **vector** is 1-dimensional sequence of items in a particular order
- the number of elements it the vector's **length**
- a vector's length can be checked with `length`
- a vector is the simplest data structure in R, hence it is called **atomic**
- a vector can store only one type of data, e.g. logical or character, thus it has a **type**
- virtually every data object is some kind of vector
- even a single data point, e.g. a `TRUE`, is actually a vector of length 1
- vectors can be created with the function `vector`: it can create any type (mode) of vector of any length
- constructors for particular types exist: `logical`, `numeric`, `character`
- when created with a constructor, all elements will be `FALSE` or `0` or `""` (empty string)
- functions exist for testing and converting types: `is.character`, `as.character`, etc.
- any vector can contain NAs

\newline

- an example of a character vector is `letters`

\newline

- vectors are most commonly built by concatenating (merging) vectors with the function `c`
```{r}
# merge two character vectors of length one
c('one', 'two')
```
- when concatenating vectors of different types, some of them will be coerced to a type that can represent all the vectors in question
```{r}
# merge a character vector with a numeric one
c('one', 1)
# the numeric is (silently!) coerced to character as a vector can only hold a single type
```

```{r}
# concatenate existing vectors (variables)
v1 <- c('one', 'two')
v2 <- c(1, 2)
v3 <- c(FALSE, TRUE)
v4 <- c(v1, v2, v3)
v4
```
- a vector can be treated as a set
- use `%in%` ro `is.element` to test if something is an element of a vector:
```{r}
'one' %in% v4
```
- a vector can contain duplicated elements
- use `anyDuplicated` to test if that is the case
- use `unique` to get unique elements of a vector

\newline

- inspect the beginning or end of a vector with `head` or `tail`, respectively (6 elements by default)
- sort a vector with `sort`
- reverse the item order with `rev`
- replicate a vector with `rep`
- create a vector of integers from i1 through i2 with the infix operator `i1 : i2`
- use `seq` for more options for numeric sequences
- use `seq_along(x)` to obtain a sequence of integers from 1 through `length(x)`

\newline

<br>

#### attributes

- a vector can have **attributes**
- an attribute is an additional feature of a vector, e.g. properties or information
- the definition of attributes is necessarily vague as they can be anything, really
- attributes are stored as a named list (more on lists below)
- all attributes of an object can be viewed with the function `attributes`
- a specific attribute can be viewed with `attr`
- a frequently encountered attribute is **names**: a vector of character strings associated with each item
- names must be unique, unless they are `NA` or `""`
- a vector that has a names attribute is a **named vector**
- see `.romans` for a named vector of Roman numerals
- names can be viewed and set with `names(x)`
- `names(x)` returns names of x
- `names(x) <- [vector of names]` sets names of x, much like assigning a variable
- an equivalent method would be to use `attr(x, "names")`
- `names` exists for convenience, as messing with names is such a common action
```{r}
# get names
names(v2)

#set names
names(v2) <- c('first', 'second')
names(v2)

# remove names
names(v2) <- NULL
names(v2)
```
- a vector can also be created with names when using `c`
- elements are given as `name = value` pairs
```{r}
v5 <- c('first' = 1, 'second' = 2, 'third' = 3)
names(v5)

# the names need not be quoted
v5 <- c(first = 1, second = 2, third = 3)
```

- we will discuss more attributes when appropriate but names will keep coming back
- the function `as.vector` will strip all atributes, leaving a bare vector

<br>

<h4>
Every vector has a type and a length. It can also have attributes.
</h4>


<br><br><br>


## II Vectorized Operations

- most operations in R are vectorized
- they are applied to each vector element separately and return a vector of the same length
```{r}
n <- 1:6

# mathematical operation
n + 1

# change case
toupper(letters)

# compare numbers
n > 3
```
- operations involving two vectors are done elementwise, by matching indices (1st with 1st, etc.)
```{r}
n1 <- 1:4
n2 <- 4:1

n1
n2
n1 + n2
n1 / n2
```
- when the lengths of the vectors involved do not match, the shorter vector is **recycled**
- this can lead to unexpected results, so watch out
```{r}
n3 <- rep(4, 4)
n4 <- 1:2

n3 + n4

l1 <- c(TRUE, FALSE)
paste(n3, l1)
```

\newline

- of course, some operations do not deal with every element of a vector separately
```{r}
s1 <- names(v5)
s1

length(s1)
```
- but many do
```{r}
nchar(s1)
```
- be careful, read the documentation, test
```{r}
l1 <- toupper(letters)
l2 <- LETTERS

identical(l1, l2)
l1 == l2

vna <- c(1, NA, 2, NA, 3, NA)

is.na(vna)
anyNA(vna)
```


<br><br><br>


## III Extraction

- **extraction**, also called **subsetting** is picking out a some elements from an object
- any language allows for extracting a single element from a sequence
- R is especially robust and flexible at it
- extraction is done with the `[` operator

\newline

- this simplest case is getting the *n*-th element of a vector: `vec[n]`
```{r}
v4[1]
```
- the number in the brackets is the index: the ordinal number of an element in the vector
- the index is not a number - it is a vector of length 1
- the index vector can be of any length
```{r}
ind <- c(1,3)
v4[ind]
```
- in fact, it can be any numeric vector at all
```{r}
ind <- c(1, 1, 3, 4, 6, 6)
v4[ind]

# invalid index yields an NA
v4[7]

# fractions are rounded down
v4[2.3]
# this should be avoided regardless
```
- if a vector is named, names can be used as indices
```{r}
v5['first']
ind <- c('first', 'second', 'first', 'first')
v5[ind]
```
- extraction can also be done with a logical vector, typically of the same length
```{r}
ind <- c(T, T, F, T, F, F)
v4[ind]
```
- a logical vector does not refer to indices and therefore its length matters
```{r}
ten <- 1:10
x <- letters[ten]
x

# too long an index produces NAs
ind <- rep(T, 11)
x[ind]

# too short an index is recycled
ind <- c(T, T, F)
x[ind]

ind <- TRUE
x[ind]
```

- a negative index will drop an item
```{r}
v4
v4[-2]
v4[-c(2,5)]
v4[c(-2, -5)]
```


<br><br><br>


## IV Replacement

- R has a **replacement** functionality, a combination of extraction and assignment
- a subset is used as a variable name in an assignment expression
- the new value is assigned to the subset in the original object
```{r}
ind <- 1:3 * 2

# extract elements
n[ind]

# replace elements
n[ind] <- NA
n
```

- assigning to a non-existent index creates a new element
```{r}
n[7] <- 15
n
```
- as well as all intervening elements
```{r}
n[10] <- 25
n
```


<br><br><br>


## V Matrix

- a matrix is a 2-dimensional data structure
- it has rows (dimension 1) and columns (dimension 2)
- a matrix is created with the function `matrix`
```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)
m
```
- by default a matrix is filled **by columns**, i.e. 1st column from 1st to last row, then 2nd column, etc.; this can be modified wyth the `byrow` argument of `matrix`
- a matrx can be transposed with the funcion `t`
```{r}
t(m)
```
- we can check the number of rows and columns with `nrow` and `ncol`, respectively

\newline

- a matrix is not a special atomic type with unique properties
- it is merely a vector with an attribute **dim**
- dim can be viewed or set with a special function `dim`
```{r}
# get dimensions
dim(m)
# the first number is the number of rows and the second is the number of columns

# the attribute can be modified as will, even removed
dim(m) <- NULL
m
# note that the matrix is dismantled in the same way is is built by default:
# column after column

# dim can be set on any vector as long as the grid size matches its length
dim(m) <- c(4,3)
m
# note that, again, the matrix is build in the default order
```
- since a matrix is basically a vector, it can hold only one type of data
- it also has a length:
```{r}
length(m)
```

<br>

#### subsetting a matrix

- when extracting from a matrix, we must account for it having two dimensions
- specify the range of each dimension and separate them with a comma:
```{r}
m[1,3]
m[1:2, 2:3]
```
- leaving one index empty will yield a whole row/column
- be aware this returns a vector, not a 1 row matrix!
```{r}
m1 <- m[1, ]
is.matrix(m1)
is.vector(m1)
```
- in order to preserve tha matrix nature of the result, we must set the `drop` argument to `FALSE`:
```{r}
m2 <- m[1, , drop = FALSE]
# mind the empty second dimension!
```
- since it is a vector, we can also subset using a single index, bypassing the dimensions:
```{r}
m[1]
m[7]
```

\newline

- a matrix can be extended by adding rows or columns with `rbind` and `cbind`, respectively:
```{r}
rbind(m, 101:103)
cbind(m, letters[1:4])
# note the coercion
```
- `c` would do no good here as it strips attributes other than names
```{r}
c(m, m)
cbind(m,m)
```

\newline

- a matrix's dimensions can be named:
```{r}
rownames(m) <- letters[1:4]
colnames(m) <- LETTERS[1:3]
m
```
- we can now use the names for subsetting
```{r}
m['a', 'B']
m['a', c('A', 'B'), drop = FALSE]
m[c('a', 'b'), c('A', 'B')]
```
- the row and column names are stored a single attribute **dimnames**
- dimanmes can be viewed or set with the function `dimnames`
- it is a list with two unnamed elements
- row and column names can be viewed or set separately with `rownames` and `colnames`
```{r}
dimnames(m)
rownames(m)
colnames(m)
```

\newline

- finally, replacement in a matrix works much the same as before
```{r}
m[2,3] <- "hello"
m
# note the conversion to character
```
- replacing to an empty index affects all items but keeps attributes
```{r}
m[] <- "goodbye"
m
```


<br><br><br>


## XI/II Array

- an **array** is an n-dimensional data structure
- it is a generalization of the matrix concept
- a matrix is basically a specific case of an nd array with n = 2
- an array is created with the function `array`
- alternatively, you can take a vector and set its `dim` to a vector longer than 2
```{r}
a1 <- array(1:12, dim = c(2, 2, 3))
a1

a2 <- 1:12
dim(a2) <- c(2, 2, 3)
a2

identical(a1, a2)
```
- extending an array, e.g. to add a matrix to a 3d array requires `abind` from the abind package
- see `iris3` for an example array

\newline

- otherwise an array behaves like a matrix
- you need to specify three dimensions when extracting
```{r}
a1[1, , ]

a1[, , 3, drop = F]
```

\newline

- arrays are not often used nowadays


<br><br><br>


## VI List

- a list is a special type of vector
- in a list any item can be of **any type** - including a list
- this is called a **recursive** object
- a list gives `TRUE` to `is.list` and `is.recursive`
- look at `state.center` and `ability.cov` for examples of lists

\newline

- a list is constructed with the function `list`
- much like `c`, `list` takes a number of objects and returns a list object
```{r}
l1 <- list(1, "one", TRUE, v5)
l1
```
- concatenate lists with `c` or `append`
- using `list` to add items to a list will take your list make it an element of a higher order list
```{r}
l2 <- list(l1, "surprize")
l2

l3 <- c(l1, "surprize")
l3
```
- turn a vector into a list with `as.list` and do the reverse with `unlist`
- a list can be named just like any other vector
- list names have restrictions just like variable names: no spaces, no special characters other than underscores or dashes (the latter never at the beginning)
- forbidden names can be used, but must be backticked
```{r}
l4 <- list(one = 1:3, two = 4:5, three = 7:8, `forbidden name` = 9)
l4
```

<br>

<h4>subsetting a list <br>
<font size = 1> it's a bit of a mess...
</font></h4>

- like any vector, a list can be extracted from
- however, the recursive structure of a list calls for special measures

\newline

- using the normal, **single bracket** returns a subset of the list
- all normal rules apply
- the result is always a list
```{r}
# subset with numeric index
l4[2]

ind <- 2:3
l4[ind]

# subset with names
l4['three']

# subset with logical vector
ind <- c(T, F, F)
l4[ind]
# the logical vector has been recycled

# it's a list
x <- l4[1]
x
is.list(x)
```

- a **double bracket** accepts only **single indices** and returns a single list item
- using a double bracket returns the **contents** of that item of the list
```{r}
# numeric
l4[[3]]

# names
l4[['forbidden name']]

# it's not a list
x <- l4[[3]]
x
is.list(x)
is.numeric(x)
```

*Note: a double bracket can also be used on a normal vector. The difference to a single bracket is that a double one discards names. It also accepts only a length 1 index. `[[` is virtually never used with atomic vectors.*

- then there is a special **dollar** operator for subsetting lists: `$`
- it only accepts bare names of list elements: `l4$one`
- it does exactly the same thing as a double bracket
```{r}
l4$one

l4$`forbidden name`

```

<br>

<h4>
A list is a kind of vector that can hold any item. <br>
Subsetting a list with `[` yields a list of any length. Using `[[` or `$` yields a single list item. <br>
</h4>


<br><br><br>


## VII R's syntactic flexibility

- there is a somewhat confusing word: **expression**
- for now, every function call is an expression
- as soon as it is executed in the console, the expression is **evaluated**
- its result is computed and possibly a value is returned
- unless it is assigned, it is passed to `print` and, well, printed
- however, assignment, like printing, is done by a function
- indeed, the value of an expression can be used by any function
- this means we can construct a complex expression where the value of one function call is used in another without assigning anything on the way
```{r}
c(1, c(4, 7))
```
- this can, obviously, be assigned rather than printed
```{r}
v6 <- c(1, c(4, 7))
```
- there are three function calls in the above expression: one to `<-` and two to `c`

- *sidenote*:

|            *the arrow operator creates a binding rather than return a value*
|            *it actually does return, but **invisibly***
|            *the value can be visualized by putting the expression in parentheses*
```{r}
(v6 <- c(1, c(4, 7)))
```
|            *this explains why we can do something like this:*
```{r}
a <- b <- 1:2
a
b
```
- piling up function calls can get unintelligible really quickly:
```{r}
head(c(paste(letters[1:3], LETTERS[8:5], c(1, 2)), 1:8), 7)
```
- still, it is very useful
```{r}
string <- 'This is a rather long character string, 
           like a sentence, it even has a period at the end.'

# get the five longest words from this sentence, in uppercase.
s <- strsplit(string, ' ')[[1]]
toupper(s[order(nchar(s), decreasing = T)][1:5])
```
- (we will learn how to deal with the commas later)


<br><br><br>


## VIII Data Frame
- a **data frame** is a 2-dimensional data structure
- it is the closest thing to a spreadsheet that exists in R
- a data frame is superficially similar to a matrix in that it has rows and columns
- **unlike** a matrix, it can hold different types of data: each column is independent
- see `iris` for an example of a data frame
- a data frame has certain dimensions (number of rows  and columns) and they can be checked
```{r}
nrow(iris)

ncol(iris)

dim(iris)
```
- test if an object is a data frame with `is.data.frame`
- a data frame has two (virtually) obligatory attributes: row names and column names
```{r}
rownames(iris)

colnames(iris)
```
- unlike `dimnames` in a matrix, `rownames` and `colnames`  are independent attributes
- both are vectors of unique character strings
- `dimnames` does still work, thoguh
```{r}
dimnames(iris)
```
- if not supplied when creating a data frame, they will be assigned default values
- `colnames` **can** be set to `NULL` and be removed
- `rownames` **can** be set to `NULL` but will be replaced by a character vector of integers

\newline

- a data frame is created with the `data.frame` function, enumerating the vectors that build columns
```{r, error=TRUE}
# different lengths of columns throw an error
data.frame(1:4, letters[1:4], c('some', 'strings', 'here'))

data.frame(1:4, letters[1:4], c('some', 'strings', 'go', 'here'))
# the automatic names are unhelpful

d1 <- data.frame(first = 1:4, 
                 second = letters[1:4], 
                 'third' = c('some', 'strings', 'go', 'here'))
d1
# note names need not be quoted
```
- matrices can be coerced to a data frame with `as.data.frame`
```{r}
as.data.frame(m)

dimnames(m) <- NULL
as.data.frame(m)
# these automatic names are better
```
- as can some lists but it the results may be surprising
```{r, error = TRUE}
l3
as.data.frame(l3)

l4
as.data.frame(l4)
```
- manually created data frames are uncommon, usually they result from importing data

\newline

- to be fair, it is possible for a column to be a list of lists rather than an atomic vector
- this is called a list column
- it is rarely used, just be aware it can happen

\newline

- a data frame is actually a special type of list with some restrictions:
    - all list elements are (usually) atomic vectors
    - all list elements are of the same length
    - the list must be named
    - the list elements are the columns of the table
- this is why `length` returns the number of columns rather than rows of a data frame
- since columns are list elements, column names are actually the names of the list elements
```{r}
colnames(iris)
names(iris)

identical(names(iris), colnames(iris))
```

\newline

- data frames are where `head` starts coming in really handy, as it displays the first rows
```{r}
head(iris)
```
- another useful function is `str`, which displays the structure of an object
```{r}
str(iris)
```
*(What's this "Factor" thing here, you ask? Hold that thought.)*

<br>

#### subsetting a data frame

- use the `[` operator and specify two dimensions
```{r}
# use numeric indices
iris[1:5, 3:4]

# use row/column names
iris[c('1', '4', '9'), c('Petal.Length', 'Species')]

# leave columns unspecified to get them all
iris[1:5, ]

# leave rows unspecified to get them all
iris[, 4:5]

# selecting a single column will yield a vector, not a data frame
iris[, 'Species']

# unless you specify that dimensions not be dropped
iris[, 'Species', drop = F]

# use logical vectors - highly unorthodox but possible
iris[c(TRUE, FALSE), c(TRUE, FALSE)]
# note the recycling

# again, selecting a single column drops dimensions
iris[, c(F, F, T, F, F)]

# unless forbidden
iris[, c(F, F, T, F, F), drop = FALSE]
```
- use the `[` operator and only specify the column(s), like with a list
```{r}
iris[4:5]
iris['Species']
# always yields a data frame, consistent with list behavior
```
- use the `[[` to get **one** column as a vector, like with a list
```{r, error=TRUE}
iris[[5]]
iris[['Species']]
# always yields a vector

iris[[5, drop = F]]
# keeping dimensions is not applicable

iris[[c('Species', 'Petal.Length')]]
# only one item

iris[[1:5]]
iris[[1:2]]
#madness
```
- use the `$` operator to extract one column as a vector, like with a list
```{r}
iris$Species
```

\newline

- so far this replicates list behavior
- now for something specific to data frames

- use a logical expression derived from the data frame to determine rows
```{r}
iris[iris$Species == 'setosa', ]
```
- what just happened?
```{r}
# 1. the iris$Species column is compared to the string 'setosa'
iris$Species == 'setosa'
# this yields a logical vector EXACTLY the same length as iris$Species, i.e. the number of rows

# 2. this vector is then used for extracting rows
ind <- iris$Species == 'setosa'
iris[ind, ]
```
- conceptually, this comes down to *get me all rows, for which the column Species has the value "setosa"*
- and since logical expressions can be combined
```{r}
iris[iris$Species == 'setosa' & iris$Sepal.Length >= 5, ]
```
- of course, this can be combined with selecting columns
```{r}
iris[iris$Species != 'virginica' & iris$Sepal.Length >= 5, 'Petal.Width']
```

<br>

#### replaceemnt in a data frame

- all the subsetting mechanics can be used for very specific replacement
```{r}
# suppose we want to omit some outliers from our calculations
# we can replace some data points with NA
iris[iris$Sepal.Length > 6 | 
       iris$Sepal.Length < 5 & 
       iris$Species == 'versicolor', 
     'Petal.Width'] <- NA
iris
```


<br>

#### operations with columns

- data frames provide a convenient way of processing data
- columns are vectors of equal length
- operations involving several columns will always be performed rowwise (no recycling)
```{r}
head(mtcars)

# compute how much horse power (hp) a car has per cylinder (cyl)
mtcars$hp / mtcars$cyl
```
- the resulting vector can easily be saved as a new column
```{r}
mtcars$hp.per.cyl <- mtcars$hp / mtcars$cyl
head(mtcars)
```
- the same can be done with the string notation with `[[`
```{r}
mtcars[['hp.per.cyl']] <- mtcars[['hp']] / mtcars[['cyl']]
```
- adding new columns is often a useful way to add information without losing data
```{r}
# rather than substituting NAs for numbers, 
# let's add a logical column that will flag certain observations as outliers
iris$outlier <- 
  ifelse(iris$Sepal.Length > 6 | 
           iris$Sepal.Length < 5 & 
           iris$Species == 'versicolor', 
         TRUE, FALSE)
iris
```
- `ifelse` is a very useful function; study it carefully

<br>

<h4>
A data frame is a special kind of list made up of atomic vectors. <br>
Extracting rows and columns is reobust. <br>
Operations on columns are easy and reliable. <br>
</h4>


<br><br><br>


## IX Factor

- remember the factor thing in iris?
```{r}
str(iris)
```
- let's have a closer look
```{r}
head(iris$Species)

str(iris$Species)
```

![a typical reaction to factors][picard]

- a **factor** is a special kind of vector, something between an integer and a character
- it is a vector of integer values
- it has a `levels` attribute
- every level is a unique character string mapped to **one** of the integer values
- see `state.region` for an example factor
- factors are created with `factor`
```{r}
ch <- rep(c('c', 'b', 'a'), 3)
f <- factor(ch)

ch

f
```
- note that levels are sorted matched to sorted values
```{r}
factor(1:3, labels = c('a', 'b', 'c'))

factor(3:1, labels = c('a', 'b', 'c'))

factor(1:3, labels = c('c', 'b', 'a'))

factor(3:1, labels = c('c', 'b', 'a'))
```
- creating factors *de novo* is rather rare
- vectors can also be converted to factors with `as.factor`
```{r}
as.factor(ch)
```
- test if a vector is a factor with `is.factor`
- it is common behavior in R to automatically convert character vectors to factors
- this is especially important when creating data frames
- `data.frame` has a `stringsAsFactors` argument just for this purpose
- beware numbers encoded as strings
```{r}
nch <- as.character(rnorm(10))

nch

as.numeric(nch)

nf <- as.factor(nch)
nf

as.numeric(nf)
```
- `as.numeric` just strips the `levels` attribute

\newline

**Wherefore the factors?**

Factors store **categorical data**. It used to be a very useful way to encode what is called discreet variables. Today much of what factors do can be don eby character vectors just as well but many procedures rely on a factor input, so they are not going away any time soon.

It may seem superfluous now but we will get to places where factors are helpful and even necessary.

I promise.

<br>

<h4>
Factors are confusing so it's all right to be confused.
</h4>


[confusedcat]: ../images/confusedcat.jpg
[picard]: ../images/picard.jpg
