---
title: "Notes 06: <br> Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(Rdlazer)
```

![lapply(c(lipstick, deodorant), drop)][droppingcat]


<br><br><br>


## I Functions recap

<i>
Here is what we have covered so far:

- functions are the active components of the language, whereas objects are passive
- a function is a block of code that can be run repeatedly
- as instance of a function running is a function call
- a function usually takes arguments
- arguments can modify the procedure within the function
- objects passed as arguments remain unchanged
- a function usually returns an output value

\newline

- R is a language where *functions are first class citizens*
- everything is done by a function
- everything can be done to a function

</i>


<br><br><br>


## II Function Environments

`If you find yourself confused by this part, consider reading the Environments vignette.`

- there are three environments that are always associated with a function
1. the enclosing environment
    - this is the environment where the function was **defined**
    - the function "remembers" the contents of its enclosing environment even when the environment no longer exists

2. the calling environment
    - this is the environment where the function is **called**
    - another, strictly technical name for it is **parent frame**
    - obviously, this can be different from call to call

3. the execution environment
    - this is an environment that is created **when** the function is called
    - each call creates a separate execution environment
    - the execution environment contains all arguments of the function and nothing else
    - the execution environment exists only for the duration of the call and is immediately deleted afterwards
    - the parent of the execution environment is the function's **enclosing** environment
*NOTE: I was wrong to say it is the calling environment during class. Like I said, the terminology is confusing. In that case it was true, however, because the function in question was both defined and called in the same (Global) environment.*


<br><br><br>


## III Defining New Functions


- you can define new function at any time
- functions are defined with the `function` function
- the arguments in the call to `function` are the **formal arguments** of the function being defined
- if formal arguments are given with a value, that becomes the default value of the argument
- the expression after the call to `function` is the **body** of the function being defined
- this is isually a compound expression (more than one expression) so it is put in braces
    - any number of expressions can consist in the body
    - any function can be called
    - variables can assigned: they will exist in the execution environment
    - functions can be defined and used within functions
- returning a value is done with a call to `return`
- a call to `return` immediately ends the function call, all subsequent expressions are ignored
```{r}
f <- function(x, y = 9L) {
  cat('x and y will be added', '\n')
  z <- x + y
  cat('the sum will be pasted to a string constant', '\n')
  ans <- paste('text', z)
  cat('the result wil be returned', '\n')
  return(ans)
  cat('I am being ignored', '\n')
  ans <- paste(ans, 'redoubled')
  return(ans)
}

# the defined function is assigned as the variable f
# f has two formal arguments, x and y; y has a default value of 9L
# during the call a variable called ans is created and returned
# returning ends the call

f(3)
```

- the call to `return` is optional
- if there is no explicit return value, the value of the last expression will be returned
```{r}
f2 <- function(x, y = 9) {
    cat('x and y will be added', '\n')
  z <- x + y
  cat('the sum will be pasted to a string constant', '\n')
  cat('the pasted text will be implicitly returned', '\n')
  paste('text', z)
}

f2(3)
```

\newline

- a function need not be assigned in order to be used
- this is called an annonymous function
```{r}
function(x, y) {z <- x + y; as.character(z)}(3, 5)

function(x, y) {x + y}(3, 5)

(function(x, y) x + y)(3, 5)
```

\newline

<i>

SIDENOTE

- during a function call arguments are evaluated as they are needed, not before
```{r, error = TRUE}
fs <- function(x, y = NULL) {
  y
  toupper(x)
}

# y is not specified, defaults to NULL
fs('text')

# y is specified as a call to stop, which signals an error and terminates the call
fs('text', stop('not so fast!', call. = FALSE))



fns <- function(x, y = NULL) {
  toupper(x)
}

fns('text')

fns('text', stop('not so fast!'))

# the error is ignored as no expression in the body refers to it
```
- this behavior is called **lazy evaluation**, as opposed to eager evaluation, and is a standard feature of R

END OF SIDENOTE
</i>


<br><br><br>


## IV Applying Functions Over Vectors

- to reiterate, a vectorized operation is one where every element of a vector is processed and a vector of the same length is returned
- not all functions perform vectorized operations

```{r}
v1 <- 1:5

# this is a vectorized operation:
v2 <- v1 < 3

# these are not
any(v2)
isTRUE(v2)
```
- quite often we want to be able to run a fuction on all elements of a vector

\newline

- R contains a set of functions specifically designed to run other functions over vectors
- the prototype of these functions is `lapply`
    - `lapply` takes a vector and a function
    - it runs the function on every element of the vector
    - it returns a **list**, where each element is the result of the functino being run on a element of the vector argument
    - it can be applied (pun uninteded) to any vector, which includes lists, which includes data frames
    - if the vector is named, the names will be transferred to the resulting list
    - again, it **always** returns a list
```{r}
v <- c('one', 'two')
vn <- setNames(v, c('item1', 'item2'))

lapply(v, identity)

lapply(vn, identity)

lapply(as.list(vn), identity)
```

\newline

- `identity` doesn't quite display the power of `lapply`
- consider the following problem: you want to compute the mean of several numeric vectors
```{r}
# get the vectors in list form
l <- as.list(iris[1:4])

# run lapply
lapply(l, mean)
```
- conceptually this is the same as running a loop
```{r}
l2 <- vector(mode = 'list', length = length(l))
for (i in seq_along(l)) {
  l2[[i]] <- mean(l[[i]])
}
names(l2) <- names(l)
l2
```
- `lapply` can be faster as the looping is done in C (the language that runs "under the hood" of R)
- `lapply` makes for neater code
- both of these are more and more true as the operation grows more complex
```{r}
ff <- function(x, text  ='text') {
  y <- mean(x)
  y <- round(y, 2)
  ans <- paste0(text, ': ', y)
  return(ans)
}

lapply(l, ff)
```

\newline

- the `X` argument in a `lapply` call, the one we iterate over, is the first formal argument of the function we apply
- other arguments can also be passed in a `lapply` call; they will be the same for every iteration

```{r}
lapply(l, ff, text = 'the mean value is')

lapply(list(iris, mtcars), head, n = 2)
```
- naming them is not strictly necesary but highly advised for clarity and safety

\newline

- finally, `lapply` accepts annonymous functions
- this is where the real flexibility shows
```{r}
# get the mean of each list element
lapply(l, mean)

# alternative with annonymous function
lapply(l, function(x) mean(x))

# get the first element of each list item
lapply(l, function(x) x[1])

# get the first and third element of each list item
lapply(l, function(x) c(x[1], x[2]))

# get the first and last elements
lapply(l, function(x) c(x[1], x[length(x)]))
```
- and it doesn't end there
```{r}
iris[, sapply(iris, is.numeric)]
```


\newline

- using the annonymous function mechanic we can also iterate over other formal arguments rather than the first
```{r}
seps <- c(',', ', ', ' - ', '_')

l[[1]]

lapply(seps, function(x) paste(l[[1]], collapse = x))
```

#### `lapply` family members

- there are other functions that build on the functionality of `lapply`

1. `sapply` does basically the same thing with two major differences:
    - it will attempt to simplify the result, *e.g.* turn a list of length 1 vectors into an atomic vector
    - it will attempt co construct names for the result based on the contents of the argument
```{r}
lapply(v, identity)

sapply(v, identity)
```
    - both of these behaviors are optional    
    - `sapply` will behave consistently but not always predictably
```{r}
# simplifying can involve coercion
ll <- list(one = '1', two = 1)
sapply(ll, identity)

# it can also be impossible
lll <- list(one = 1:2, two = 1)
sapply(lll, identity)
```

2. `vapply` is `lapply` with safety measures:
    - it will also attempt to simplify and create names
    - you must specify the output you expect from `FUN`, *e.g.* "character of length 1"
```{r, error = TRUE}
vapply(ll, identity, character(1))

vapply(v, identity, character(1))
```
    - `vapply` is recommended over `sapply` in programming (calling it within a function)

3. `mapply` is a "multivariate" version of `sapply`
    - it can iterate over more than one argument
```{r}
mapply(paste, l, collapse = seps)
```
    - the variable arguments are passed (named) in `...`
    - arguments that are to be constant across iterations are passed in `MoreArgs` as a list
    - mind the difference in syntax:
    `sapply(X, FUN)` (vector first) vs. `mapply(FUN, ..., MoreArgs)` (function first)

4. `replicate` is a little different
    - it does not iterate over a vecotr
    - it will run an operation a set number of times and simplify the result
```{r}
replicate(5, 1:4)
```
    - this is handy for simulating random processes
```{r}
# create 5 random samples (10 items each) of a normal distribution
replicate(5, rnorm(10))

# compute mean of a random sample (10k items)
replicate(100, {
  s <- rnorm(1e+4)
  mean(s)
})

# capture the means and compute their mean
means <- replicate(100, {
  s <- rnorm(1e+4)
  sm <- mean(s)
})

head(means)

mean(means)
```
- note that when we assign the result of a random operation to a variable, we assing the **value** of that one call rather than an unevaluated call itself
- this means the call is **not** repeated when we call the assigned variable:
```{r}
# random pertumation
sample(6)

# another random permutation
sample(6)

# assigned random permutation ("captured")
(p <- sample(6))

# the permutation remains "captured"
p

# p stores the value returned by the call to sample, not the function call itself
```

\newline

*Check out the `Vectorize` function too.*

<h4>
`lapply` is a very powerful and flexible tool <br>
`sapply` can return vectors rather than lists <br>
`sapply` results can be surprising <br>
`mapply` can iterate over any number of arguments <br>
`replicate` can be useful for simulations <br>

</h4>

[droppingcat]: ../images/droppingcat_cropped.gif
