---
title: "Character Strings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Notes 12: Character Strings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


<br><br><br>


## Character String Basics

- a character string is a specific data type
- a string is composed of characters: 
    - letters, lower- upper case
    - digits
    - punctuation marks and other special characters (`,;:@#$%` etc.)
    - whitespace characters: space, tabulator, line break
- characters are contained within quote marks, single or double
- the quote marks are not treated literally but have a special meaning: they limit the contents a character string
- the quote marks are a **metacharacter** (more on metacharactes below)
- a string is "closed" by the same type of quote mark it was "opened" with
- a single quote mark within a double-quoted string will be treated literally
- check the number of characters in a string with `nchar`
- character case can be changed with `tolower` and `toupper`
- elements of a character vector can be changed with `replace`
```{r}
v <- names(precip)[3:8]
v

replace(v, list = 1:3, c('replacement 1', 'replacement 2'))
```

\newline

#### metacharacters

- **metacharacters** are characters whose meaning is not literal
- e.g. a quote mark in not a quote mark, it opens and closes a sting
- special meaning can be stripped from a metacharacter so that it will be treated literally
- this is done with an **escape character** and is called **escaping**
- the escape character in R (and other languages as well) is the backslash: `\`
- and so, a literal quote mark can be put in a string with `\"`
- *note: in a string delimited by `"`, `'` is treated literally, and vice versa*

\newline

- some characters cannot be put into a string directly and they must be encoded with metacharacters
- in such cases the the escape character bestows meaning rather than stripping it
- and so, a tabulator is encoded by `\t` and a line break by `\n`
- a backslash will also require escaping, so `\\`

\newline

## Combining And Splitting Strings

- strings are concatenated (combined) with the function `paste`
- any number of character strings and character vectors can be concatenated at once but mind the vector recycling mechanics
- the `collapse` argument allows `paste` to combine a character vector into one string
```{r}
paste('some', 'text')

paste(c('some', 'more'), 'text')

paste(c('some', 'text'), collapse = '_')
```
- `paste0` is a fast wrapper with no separator

<br>

- the `substr` function can isolate a range of a string, beginning and ending at specified positions
```{r}
substr('some text', start = 3, stop = 8)
```

\newline

- a string can also be split into several strings with `strsplit`
- the `split` argument specifies the character(s) that the separation will be done on
- the separating characters are removed from the result
```{r}
strsplit('even_more_text', split = '_')
```
- note that a list is returned; the length of the list is equal to the that of `x`
```{r}
strsplit(v, split = ' ')
```
- we can now easily count the number of per sentence in a piece of text:
```{r, out.width='.8\\textwidth'}
text <- "In the Lenin Barracks in Barcelona, the day before I joined the militia, I saw an Italian militiaman standing in front of the officers' table.

He was a tough-looking youth of twenty-five or six, with reddish-yellow hair and powerful shoulders. His peaked leather cap was pulled fiercely over one eye. He was standing in profile to me, his chin on his breast, gazing with a puzzled frown at a map which one of the officers had open on the table. Something in his face deeply moved me. It was the face of a man who would commit murder and throw away his life for a friend--the kind efface you would expect in an Anarchist, though as likely as not he was a Communist. There were both candour and ferocity in it; also the pathetic reverence that illiterate people have for their supposed superiors. Obviously he could not make head or tail of the map; obviously he regarded map-reading as a stupendous intellectual feat. I hardly know why, but I have seldom seen anyone--any man, I mean--to whom I have taken such an immediate liking. While they were talking round the table some remark brought it out that I was a foreigner."

# split text on period followed by any character
sentences <- strsplit(text, split = '\\..')[[1]]
sentences

sapply(strsplit(sentences, ' '), length)
```
- note that unless `fixed` is `TRUE`, `split` is interpreted as a regular expression (see below)


<br><br><br>


## Matching Strings

- it is common to test whether a string contains a certain substring
- of course, this can be done on a vector
- `startsWith` and `endsWith` check for matches at the edges of strings
```{r}
vv <- c('word 1', 'word 2', 'word3', 'Word 4', 'Words', 'word')

startsWith(vv, 'word')

endsWith(vv, 's')
```
- the `grep` family of functions provedes very useful funcionalities
```{r}
# grep returns indices of matching vector elements
grep('word', vv)

# it can also return the matching items
grep('word', vv, value = TRUE)

# or the non-matching ones
grep('word', vv, value = TRUE, invert = TRUE)

# character case can be ignored
grep('word', vv, value = TRUE, ignore.case = TRUE)

# grepl returns a logical vector based on the presence of absence of a match
grepl('word', vv)

# both functinos can ignore case
grepl('word', vv, ignore.case = TRUE)
```


<br><br><br>


## Regular Expressions

The first argument fo `grep` and `grepl` specifies the pattern that is matched to elements of a caracter vector. Normally every character is matched exactly but it is possible to loosen the matching process. **Regular expressions** (often abbreviated as **regex**) are a way of ambiguating a the search pattern so that non-exact matches are identified. Matches can occur on alternative characters, repeated characters and more.

The ambiguation is achieved by using wildcards, which are put into the pattern as metacharacters. The most often used wildcards are:

- `.` matches any character
- `[]` matches any character within the brackets
    - `[abcde]` matches a, b, c, d or e
    - `[a-e]` also matches a, b, c, d or e
    - `[a-z]` matches any lower case letter
    - `[4-9]` matches any digit from 4 to 9
    - `[49]` matches either a 4 or a 9
    - `[0-9A-Z]` matches any digit or any upper case letter
- `[^]` mathes any character **other** than the ones within the brackets following a `^`
- `^` stands for the beginning of the string
- `$` stands for the end of the string
- `*` the previous character is repeated any number of times
- `+` the previous character is repeated at least 1 time
- `?` the previous character is repeated 0 or 1 time
- `{2,4}` the previous character is repeated at least 2 and at most 4 times
- `|` separates alternative expressions and allows a match of any
- `()` parenetheses delimit subexpressions, which is used for **backreference** (see below)

There are other ways of specifying character classes, e.g.:

- `\w` any letter, digit, or underscore
- `\d` any digit
- `\s` any whitespace character

Whenever a wildcard character is to be taken literally, it must be escaped. Importantly, in R the backslash alone is a literal character so it must be escaped to turn int into the escape character. And so `\.` means "backslash-period" but `\\.` means "any character".

Some examples:

- `^A` matches anything beginning with A, including a single A
- `^A.` a string beginning with A followed by any character
- `^A.$` matches a string containing A followed by any one character and nothing else
- `A*$` mathces a string ending with any number of As

\newline

- ` ` matches any string containing a space
- ` u` matches a space followed by a u
- ` T?u{1,3}$` matches a space followed by an optional T, followed by one to three us but *only* at the end of a string

\newline

- `grey` matches only grey
- `grey|gray` matches both gray and grey (US vs UK spelling)
- `gr[ae]y` is equivalent to the above - does *not* match graey!
- `[Ee]arth` matches Earth and earth
- `[Ee]arths?` also matches Earths and earths

\newline

- `^[A-Z].*\\.$` matches a string that begins with a capital letter, ends with a period and contains any number of any characters in between - a typical declarative sentence

- `^[A-Z].*, .*\\.$` matches a sentence that contains a comma followed by a space anywhere within, which indicates a compond sentence

- `^[A-Z].*, .*[.?]` the sentence can now also be a question; note that characters within brackets are taken literally and need not be escaped

\newline

More examples:
```{r}
vv

grep('^w', vv, value = TRUE)

grep('^[wW]', vv, value = TRUE)

grep('^Ws$', vv, value = TRUE)

grep('^W.s$', vv, value = TRUE)

grep('^W.+s$', vv, value = TRUE)

grep('^W.+s?$', vv, value = TRUE)

grep(' \\d$', vv, value = T)

grep(' *\\d$', vv, value = T)
```

<br>

#### replacing patterns

The `sub` function searches a string for matches to a regular expression just like `grep` does. However, rather than reporting matches, it replaces them with a replacement string. The replacement is a literal string, not a regex.
```{r}
days <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday', 'Sunday')

grep('day', days, value = TRUE)

sub('day', 'DAY', days)
```

`sub` only acts on the first occurrence of the pattern. To replace all accurrences use `gsub`.
```{r}
sub('e', '.E.', 'Wednesday')

gsub('e', '.E.', 'Wednesday')
```

<br>

#### backtracing

Characters in a regular expression can be grouped using parentheses: `(Wednes)(day)`. The ability to recall these groups individually is called **backtracing**. Backtracing allows for preserving a partial match during substitution, whereas normally the entire match is replaced.
```{r}
sub('day', 'DAY', 'Wednesday')

sub('.*day', 'DAY', 'Wednesday')

sub('(.*)(day)', 'DAY', 'Wednesday')

sub('(.*)(day)', '\\1\\2', 'Wednesday')

sub('(.*)(day)', '\\1', 'Wednesday')

sub('(.*)(day)', '\\1DAY', 'Wednesday')
```

<h4>
Regular expressions are a powerful tool for searching patterns within text. <br>
They also allow for full and partial replacing of the mathcing patterns.
</h4>
