---
title: "Notes 05: <br> Packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

![a package][kittens]


<br><br><br>


## I Packages

- a **package** is a collection of files containing code, *i.e.* primarily functions
- a package can also contain data, either as raw files or as code that generates the data
- some packages contain only data - this allows the data to be neatly documented
- a package is a common way of distributing code, ensuring it will be readily usable
- in other languages (*e.g.* Python, or C++) this is usually called a library



#### the library

- before they can be used, packages have to be **installed**
- install packages with `install.packages`
- this downloads the specified packages from a **repository** and installs them
- the default repository is **CRAN**: the entity that keeps, maintains, and delevops the R language
- there are other repositories, *e.g.* BioConductor, Github
- non-CRAN repositories can have other ways to install packages
- all installed packages are kept in a directory called the **library**
- list all installed packages with `.packages(all.available = TRUE)`
- for more comprehensive view use `installed.packages`
- uninstall packages with `remove.packages`
- installing a package does **not** mean you can access its contents, not quite yet

#### packages in action

- a package has a **namespace**, a collection of names to which values are assigned
- objects that are made accessible to the user are said to be **exported**
- a namespace also contains names of objects that are **imported** from other packages but that is another story
- a package can be **attached** to a session with `library`
- this causes the contents of the namespace to be accessible
```{r, error = TRUE}
hint('0.1.2')

library(Rdlazer)
hint('0.1.2')
```

\newline

- note that attaching `Rdlazer` causes **masking** of two objects from `datasets`
- masking means that two objects exist under one name
- by defualt the name will access the object that was attached later
- you can refer to them unambiguously with the special operator `::`
```{r}
head(mtcars)

head(Rdlazer::mtcars)

head(datasets::mtcars)
```
- the `::` operator can also be used to access namespaces of packages that have been installed but not attached
```{r, error = TRUE}
head(diamonds)

head(ggplot2::diamonds)
```

\newline

- **detach** a package from the current session with `detach("package:packagename", unload = TRUE)`
- the currently attached packages can be viewed with the function `.packages`
```{r}
# .packages returns invisibly
(.packages())
```
- another way is to view the search list with `search`; check the Environments vignette for an explanation

[kittens]: ../images/kittenspackage.png
