---
title: "Notes For Part Two"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Notes For Part Two}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

![a tour of R][confusedcat]

# Data Structures And *Language Properties*
<font size = 1> **with some function vocabulary** </font>





## I Vector

- a **vector** is 1-dimensional sequence of items in a particular order
- the number of elements it the vector's **length**
- a vector's length can be checked with `length`
- a vector is the simplest data structure in R, hence it is called **atomic**
- a vector can store only one type of data, e.g. logical or character, thus it has a **type**
- virtually every data object is some kind of vector
- even a single data point, e.g. a `TRUE`, is actually a vector of length 1
- vectors can be created with the function `vector`: it can create any type (mode) of vector of any length
- constructors for particular types exist: `logical`, `numeric`, `character`
- when created with a constructor, all elements will be `FALSE` or `0` or `""` (empty string)
- functions exist for testing and converting types: `is.character`, `as.character`, etc.
- any vector can contain NAs

\newline

- an example of a character vector is `letters`

\newline

- vectors are most commonly built by concatenating (merging) vectors with the function `c`
```{r}
# merge two character vectors of length one
c('one', 'two')
```
- when concatenating vectors of different types, some of them will be coerced to a type that can represent all the vectors in question
```{r}
# merge a character vector with a numeric one
c('one', 1)
# the numeric is (silently!) coerced to character as a vector can only hold a single type
```

```{r}
# concatenate existing vectors (variables)
v1 <- c('one', 'two')
v2 <- c(1, 2)
v3 <- c(FALSE, TRUE)
v4 <- c(v1, v2, v3)
v4
```
- a vector can be treated as a set
- use `%in%` ro `is.element` to test if something is an element of a vector:
```{r}
'one' %in% v4
```
- a vector can contain duplicated elements
- use `anyDuplicated` to test if that is the case
- use `unique` to get unique elements of a vector

\newline

- inspect the beginning or end of a vector with `head` or `tail`, respectively (6 elements by default)
- sort a vector with `sort`
- reverse the item order with `rev`
- replicate a vector with `rep`
- create a vector of integers from i1 through i2 with the infix operator `i1 : i2`
- use `seq` for more options for numeric sequences
- use `seq_along(x)` to obtain a sequence of integers from 1 through `length(x)`

\newline

- a vector can have **attributes**
- an attribute is an additional feature of a vector, e.g. properties or information
- the definition of attributes is necessarily vague as they can be anything, really
- all attributes of an object can be viewed with the function `attributes`
- a specific attribute can be viewed with `attr`
- a frequently encountered attribute is **names**: a vector of character strings associated with each item
- names must be unique, unless they are `NA` or `""`
- a vector that has a names attribute is a **named vector**
- see `.romans` for a named vector of Roman numerals
- names can be viewed and set with `names(x)`
- `names(x)` returns names of x
- `names(x) <- [vector of names]` sets names of x, much like assigning a variable
- an equivalent method would be to use `attr(x, "names")`
- `names` exists for convenience, as messing with names is such a common action
```{r}
# get names
names(v2)

#set names
names(v2) <- c('first', 'second')
names(v2)

# remove names
names(v2) <- NULL
names(v2)
```
- a vector can also be created with names when using `c`
- elements are given as `name = value` pairs
```{r}
v5 <- c('first' = 1, 'second' = 2, 'third' = 3)
names(v5)

# the names need not be quoted
v5 <- c(first = 1, second = 2, third = 3)
```

- we will discuss more attributes when appropriate but names will keep coming back
- the function `as.vector` will strip all atributes, leaving a bare vector

#### Every vector has a type and a length. It can also have attributes.





## II *Vectorized Operations*

- most operations in R are vectorized
- they are applied to each vector element separately and return a vector of the same length
```{r}
n <- 1:6

# mathematical operation
n + 1

# change case
toupper(letters)

# compare numbers
n > 3
```
- operations involving two vectors are done elementwise, by matching indices (1st with 1st, etc.)
```{r}
n1 <- 1:4
n2 <- 4:1

n1
n2
n1 + n2
n1 / n2
```
- when the lengths of the vectors involved do not match, the shorter vector is **recycled**
- this can lead to unexpected results, so watch out
```{r}
n3 <- rep(4, 4)
n4 <- 1:2

n3 + n4

l1 <- c(TRUE, FALSE)
paste(n3, l1)
```

\newline

- of course, some operations do not deal with every element of a vector separately
```{r}
s1 <- names(v5)
s1

length(s1)
```
- but many do
```{r}
nchar(s1)
```
- be careful, read the documentation, test
```{r}
l1 <- toupper(letters)
l2 <- LETTERS

identical(l1, l2)
l1 == l2

vna <- c(1, NA, 2, NA, 3, NA)

is.na(vna)
anyNA(vna)
```





## III *Extraction*

- **extraction**, also called **subsetting** is picking out a some elements from an object
- any language allows for extracting a single element from a sequence
- R is especially robust and flexible at it
- extraction is done with the `[` operator

\newline

- this simplest case is getting the *n*-th element of a vector: `vec[n]`
```{r}
v4[1]
```
- the number in the brackets is the index: the ordinal number of an element in the vector
- the index is not a number - it is a vector of length 1
- the index vector can be of any length
```{r}
ind <- c(1,3)
v4[ind]
```
- in fact, it can be any numeric vector at all
```{r}
ind <- c(1, 1, 3, 4, 6, 6)
v4[ind]

# invalid index yields an NA
v4[7]

# fractions are rounded down
v4[2.3]
# this should be avoided regardless
```
- if a vector is named, names can be used as indices
```{r}
v5['first']
ind <- c('first', 'second', 'first', 'first')
v5[ind]
```
- extraction can also be done with a logical vector, typically of the same length
```{r}
ind <- c(T, T, F, T, F, F)
v4[ind]
```
- a logical vector does not refer to indices and therefore its length matters
```{r}
ten <- 1:10
x <- letters[ten]
x

# too long an index produces NAs
ind <- rep(T, 11)
x[ind]

# too short an index is recycled
ind <- c(T, T, F)
x[ind]

ind <- TRUE
x[ind]
```
<font color = 'blue'>
- a negative index will drop an item
```{r}
v4
v4[-2]
v4[-c(2,5)]
v4[c(-2, -5)]
```
</font>




## IV *Replacement*

- R has a **replacement** functionality, a combination of extraction and assignment
- a subset is used as a variable name in an assignment expression
- the new value is assigned to the subset in the original object
```{r}
ind <- 1:3 * 2

# extract elements
n[ind]

# replace elements
n[ind] <- NA
n
```
<font color = 'blue'>
- assigning to a non-existent index creates a new element
```{r}
n[7] <- 15
n
```
- as well as all intervening elements
```{r}
n[10] <- 25
n
```
</font>




## V Matrix

- a matrix is a 2-dimensional data structure
- it has rows (dimension 1) and columns (dimension 2)
- a matrix is created with the function `matrix`
```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)
m
```
- by default a matrix is filled **by columns**, i.e. 1st column from 1st to last row, then 2nd column, etc.; this can be modified wyth the `byrow` argument of `matrix`
- a matrx can be transposed with the funcion `t`
```{r}
t(m)
```
- we can check the number of rows and columns with `nrow` and `ncol`, respectively

\newline

- a matrix is not a special atomic type with unique properties
- it is merely a vector with an attribute **dim**
- dim can be viewed or set with a special function `dim`
```{r}
# get dimensions
dim(m)
# the first number is the number of rows and the second is the number of columns

# the attribute can be modified as will, even removed
dim(m) <- NULL
m
# note that the matrix is dismantled in the same way is is built by default:
# column after column

# dim can be set on any vector as long as the grid size matches its length
dim(m) <- c(4,3)
m
# note that, again, the matrix is build in the default order
```
- since a matrix is basically a vector, it can hold only one type of data
- it also has a length:
```{r}
length(m)
```

#### subsetting a matrix
- when extracting from a matrix, we must account for it having two dimensions
- specify the range of each dimension and separate them with a comma:
```{r}
m[1,3]
m[1:2, 2:3]
```
- leaving one index empty will yield a whole row/column
- be aware this returns a vector, not a 1 row matrix!
```{r}
m1 <- m[1, ]
is.matrix(m1)
is.vector(m1)
```
- in order to preserve tha matrix nature of the result, we must set the `drop` argument to FALSE:
```{r}
m2 <- m[1, , drop = FALSE]
# mind the empty second dimension!
```
- since it is a vector, we can also subset using a single index, bypassing the dimensions:
```{r}
m[1]
m[7]
```

\newline

- a matrix can be extended by adding rows or columns with `rbind` and `cbind`, respectively:
```{r}
rbind(m, 101:103)
cbind(m, letters[1:4])
# note the coercion
```
- `c` would do no good here as it strips attributes other than names
```{r}
c(m, m)
cbind(m,m)
```

\newline

- a matrix's dimensions can be named:
```{r}
rownames(m) <- letters[1:4]
colnames(m) <- LETTERS[1:3]
m
```
- we can now use the names for subsetting
```{r}
m['a', 'B']
m['a', c('A', 'B'), drop = FALSE]
m[c('a', 'b'), c('A', 'B')]
```
- the row and column names are stored a single attribute **dimnames**
- dimanmes can be viewed or set with the function `dimnames`
- it is structured as a list (more on lists below)

\newline

- finally, replacement in a matrix works much the same as before
```{r}
m[2,3] <- "hello"
m
# note the conversion to character
```
- replacing to an empty index affects all items but keeps attributes
```{r}
m[] <- "goodbye"
m
```





## VI List

- a list is a special type of vector
- in a list any item can be of **any type** - including a list
- this is called a **recursive** object
- a list gives `TRUE` to `is.list` and `is.recursive`
- look at `state.center` and `ability.cov` for examples of lists

\newline

- a list is constructed with the function `list`
- much like `c`, `list` takes a number of objects and returns a list object
```{r}
l1 <- list(1, "one", TRUE, v5)
l1
```
- concatenate lists with `c` or `append`
- using `list` to add items to a list will take your list make it an element of a higher order list
```{r}
l2 <- list(l1, "surprize")
l2

l3 <- c(l1, "surprize")
l3
```
- turn a vector into a list with `as.list` and do the reverse with `unlist`
- a list can be named just like any other vector
- list names have restrictions just like variable names: no spaces, no special characters other than underscores or dashes (the latter never at the beginning)
- forbidden names can be used, but must be backticked
```{r}
l4 <- list(one = 1:3, two = 4:5, three = 7:8, `forbidden name` = 9)
l4
```

<h4>subsetting a list <br>
<font size = 1> it's a bit of a mess... </font>
</h4>

- like any vector, a list can be extracted from
- however, the recursive structure of a list calls for special measures

\newline

- using the normal, **single bracket** returns a subset of the list
- all normal rules apply
- the result is always a list
```{r}
# subset with numeric index
l4[2]

ind <- 2:3
l4[ind]

# subset with names
l4['three']

# subset with logical vector
ind <- c(T, F, F)
l4[ind]
# the logical vector has been recycled

# it's a list
x <- l4[1]
x
is.list(x)
```

- a **double bracket** accepts only **single indices** and returns a single list item
- using a double bracket returns the **contents** of that item of the list
```{r}
# numeric
l4[[3]]

# names
l4[['forbidden name']]

# it's not a list
x <- l4[[3]]
x
is.list(x)
is.numeric(x)
```

*Note: a double bracket can also be used on a normal vector. The difference to a single bracket is that a double one discards names. It also accepts only a length 1 index. `[[` is virtually never used with atomic vectors.*

- then there is a special **dollar** operator for subsetting lists: `$`
- it only accepts bare names of list elements: `l4$one`
- it does exactly the same thing as a double bracket
```{r}
l4$one

l4$`forbidden name`

```

<h4> A list is a kind of vector that can hold any item. <br>
Subsetting a list with `[` yields a list of any length. Using `[[` or `$` yields a single list item.
<h4>





## VII *R's syntactic flexibility*

- there is a somewhat confusing word: **expression**
- for now, every function call is an expression
- as soon as it is executed in the console, the expression is **evaluated**
- its result is computed and possibly a value is returned
- unless it is assigned, it is passed to `print` and, well, printed
- however, assignment, like printing, is done by a function
- indeed, the value of an expression can be used by any function
- this means we can construct a complex expression where the value of one function call is used in another without assigning anything on the way
```{r}
c(1, c(4, 7))
```
- this can, obviously, be assigned rather than printed
```{r}
v6 <- c(1, c(4, 7))
```
- there are three function calls in the above expression: one to `<-` and two to `c`

- *sidenote*:

|            *the arrow operator creates a binding rather than return a value*
|            *it actually does return, but **invisibly***
|            *the value can be visualized by putting the expression in parentheses*
```{r}
(v6 <- c(1, c(4, 7)))
```
|            *this explains why we can do something like this:*
```{r}
a <- b <- 1:2
a
b
```
- piling up function calls can get unintelligible really quickly:
```{r}
head(c(paste(letters[1:3], LETTERS[8:5], c(1, 2)), 1:8), 7)
```
- still, it is very useful
```{r}
string <- 'This is a rather long character string, 
           like a sentence, it even has a period at the end.'

# get the five longest words from this sentence, in uppercase.
s <- strsplit(string, ' ')[[1]]
toupper(s[order(nchar(s), decreasing = T)][1:5])
```
- (we will learn how to deal with the commas later)







```{r forlater, eval=FALSE, include=FALSE}
# ## VIII Data Frame
# - the data frame is pehaps the most often used data structure
```

[confusedcat]: ../images/confused.jpg
